# Analysis of major storm and weather events in 
# the USA from 1950--2011

## Synopsis

The U.S. National Oceanic and Atmospheric Administration's (NOAA) [storm database](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)
was used to compare outcomes from storm and severe weather event categories.
Analyzed data are from 1950--2011.
Outcomes analyzed are the number of fatalities and injuries and the dollar
amount of property damage.
Event categories are for example winter events, hurricane events, wind
events, heat, tornado, flood events. The second name 'events' is
indicator that multiple events with different names but similar behaviour
are grouped as name of the dominant event + event name. Wind type is exception.

The results of analysis are shown in tables and plots. The key of analysis
is find the most influential event types for fatalities, injuries,
property and crop damage.


## Data Processing

### 1. Read data file

The data file (bz2) is in the working directory.
Read the bz2 file as csv, select columns by col_names and convert data
frame to data.table

```{r}

library(data.table)
library(plyr)
library(gridExtra)
library(ggplot2)

col_names <- c("EVTYPE","FATALITIES","INJURIES","PROPDMG","PROPDMGEXP","CROPDMG","CROPDMGEXP")
# data <- data.table(read.csv(bzfile("./repdata_data_StormData.csv.bz2"))[,col_names])
data <- fread("./Storm_data.csv")

```
### 2. Lower case the column names and evtype, show how much are event names repeating

```{r}

# Change the column names to lower case
setnames(data, tolower(names(data)))

# Change all event names in evtype column to lower case
data[,evtype := tolower(evtype)][1:5]

# How much are the names repetitive in the evtype column? 
arrange(data[, .N, by = evtype], desc(N))[1:40]

# How much are the specific event names repetitive?
arrange(data[grep("wind", evtype), .N, by = evtype], desc(N))[1:20]

arrange(data[grep("flood", evtype), .N, by = evtype], desc(N))[1:20]

```
### 3. Join the most influential events within evtype column

I try to join similar names of the most influential source of events and in
addition join the names of events with similar behaviour but different (not similar) names. 
The first name of every string in the events list will be the representing name in evtype 
column for all names in the string. This is done through gsub function.

```{r}

events <- list("heavy rain|rain", "winter events|winter|frost|cold|snow|freez|ice|blizzard", 
    "hurricane events|hurricane|typhoon|tropical|waterspout", "heat|warm", "hail", 
    "flood events|flood|surf|surge", "wind events|wind", "tornado", "rip", "fog", "fire")

# This is a very complex line that deal with data, grep every element of list events within data 
# column evtype and assing this grep variables to evtype as the first name from events string
# through gsub function. The solution is within the data because data.table package (different than data.frame).
tmp <- lapply(events, function(x) data[grep(x, evtype), evtype := gsub("\\|+[a-z]+", "", x)])

data

data[grep("wind", evtype), .N, by = evtype]

data[grep("flood", evtype), .N, by = evtype]

```

### 4. The most harmful events with respect to population health

Subset sum of injuries and fatalities by evtype and show ordered subset data. 

```{r}

harm_sum <- data[, list(fatal = sum(fatalities), injur = sum(injuries)), by = evtype]
fatal <- harm_sum[order(-fatal), fatal, evtype][1:15]
fatal
injur <- harm_sum[order(-injur), injur, evtype][1:15]
injur

```
### 5. The events of greatest economic consequences 

#### 5.1 The property damage

```{r}

# create data table for merging propdmgexp
prop_token <- data.table(propdmgexp = c("K", "M", "B"), dollars = as.numeric(c("1000", "10e5", "10e8")))

# merge and convert propdmgexp tokens into dollars
prop_merge <- merge(prop_token, data, by = "propdmgexp")

prop_merge

#####################################################################################
#  ALTERNATIVE 1

# prop_merge <- merge(setkey(prop_token, propdmgexp), setkey(data, propdmgexp))

# ALTERNATIVE 2 (in one line) 

# prop_merge_alt<- data[propdmgexp %like% "K|M|B", list(evtype, prop = ifelse(propdmgexp == "B", propdmg * 1e+09, ifelse(propdmgexp == 
#    "M", propdmg * 1e+06, ifelse(propdmgexp == "K", propdmg * 1000, na.omit(propdmg)))))]

# AND SUM BY EVTYPE

# prop_merge_alt[, list(prop = sum(prop)), by = evtype]
#####################################################################################

prop_sum <- prop_merge[, list(prop = sum(propdmg * dollars)), by = evtype]

######################################################################################
#  ALTERNATIVE

# prop_sum <- ddply(prop_merge[, prop := propdmg * dollars], .(evtype), summarize, prop = sum(prop))
######################################################################################

prop <- prop_sum[order(-prop)][1:15]
prop

```

#### 5.2 The crop damage

```{r}

crop_token <- data.table(cropdmgexp = c("K", "M", "B"), dollars = as.numeric(c("1000", "10e5", "10e8")))
crop_merge <- merge(crop_token, data, by = "cropdmgexp")
crop_sum <- crop_merge[, list(crop = sum(cropdmg * dollars)), by = evtype]
crop <- crop_sum[order(-crop)][1:15]
crop

```
## Results

```{r fig.width = 15, fig.height = 7}

f <- ggplot(fatal, aes(x = reorder(evtype, fatal), y = fatal)) + geom_bar(stat = "identity") + coord_flip() +
theme(axis.text.y = element_text(size=20,color="darkred"), text = element_text(size=19), axis.title.y = element_text(vjust = 1.0), 
      axis.title.x = element_text(vjust = -0.6), plot.title = element_text(size = 19, vjust = 2)) +
labs(y = "Deaths", x = "Weather events") + ggtitle("Weather fatalities in the USA for 1950-2011")

i <- ggplot(injur, aes(x = reorder(evtype, injur), y = injur)) + geom_bar(stat = "identity") + coord_flip() +
theme(axis.text.y = element_text(size=20,color="darkred"), text = element_text(size=19), axis.title.y = element_text(vjust = 1.0), 
      axis.title.x = element_text(vjust = -0.6), plot.title = element_text(size = 19, vjust = 2)) +
labs(y = "Injuries", x = "") + ggtitle("Weather injuries in the USA for 1950-2011")

grid.arrange(f, i, ncol = 2)

p <- ggplot(prop, aes(x = reorder(evtype, prop), y = prop)) + geom_bar(stat = "identity") + coord_flip() +
theme(axis.text.y = element_text(size=20,color="darkred"), axis.text.x = element_text(size=15, color="darkred"), text = element_text(size=19), 
      axis.title.y = element_text(vjust = 1.0), axis.title.x = element_text(vjust = -0.6), plot.title = element_text(size = 19, vjust = 2)) +
labs(y = "Property damage in Dollars", x = "Weather events") + ggtitle("Property damage - USA for 1950-2011")

c <- ggplot(crop, aes(x = reorder(evtype, crop), y = crop)) + geom_bar(stat = "identity") + coord_flip() +
theme(axis.text.y = element_text(size=20,color="darkred"), axis.text.x = element_text(size=15, color="darkred"), text = element_text(size=19), 
      axis.title.y = element_text(vjust = 1.0), axis.title.x = element_text(vjust = -0.6), plot.title = element_text(size = 19, vjust = 2)) +
labs(y = "Crop damage in Dollars", x = "") + ggtitle("Crop damage - USA for 1950-2011")

grid.arrange(p, c, ncol = 2)


```
